---
title: "Using the Adult Weight Change Model"
author: "Rodrigo Zepeda Tello & Dalia Camacho García Formentí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adult Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: Referencias.bib
---

```{r, echo = FALSE, message=FALSE, warning = FALSE}
require("bw")
require("ggplot2")
```

> In this vignette we explain how to use the model for adults in R; we develop and explain the equations involved both for casual and advanced readers. 

#Contents 

+ [Usage in R](#user)
+ [Basic model explanation](#basic)
+ [Advanced model explanation](#advanced)


#Usage in R {#user}

##Inputs
The main inputs for the body weight change model in adults are:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `bw`  |  Body Weight (kg)  | No | - |
|  `ht`  |  Height (m) | No | - |
|  `age` |  Age (yrs) | No | - |
|  `sex` |  Either `'male'` or `'female'` | No | - |

As an example consider a `40` year old `'female'` weighting `80` kg with a height of `1.8` metres: 

```{r}
female_model1 <- adult_weight(bw = 80, ht = 1.8, age = 40, sex = "female")
```

Furthermore, the model allows the user to input daily changes in sodium and energy intake compared to the baseline consumption (for example, if doing a diet which reduces `-100` kcals from previous consumption). 

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `EIchange`  | Energy intake difference from baseline (kcal) | Yes | No energy change |
|  `NAchange`  | Sodium intake difference from baseline (kcal) | Yes | No sodium change |

Note that this `-100` kcals is not cummulative. For example, if an individual consumes daily `2000` kcals a `EIchange = -100` means that in day 1 the individual consumes `1900` kcals; in day 2 the individual consumes `1900` kcals; in day 3 the individual consumes `1900` kcals, etc. 

For example, this female can reduce her energy consumption by `-250` kcals and her sodium intake by `20` mg:

```{r}
female_model2 <- adult_weight(bw = 80, ht = 1.8, age = 40, sex = "female", 
                      EIchange = rep(-250, 365), NAchange = rep(-20, 365))
```

If no `NAchange` is inputed, the model assumes a `0` change:
```{r}
female_model3 <- adult_weight(bw = 80, ht = 1.8, age = 40, sex = "female", 
                      EIchange = rep(-250, 365))
```

The same is true for `EIchange`. 

Note that in the examples above, `EIchange = rep(-250, 365)` is inputed as a vector with each time step `dt` (default, `dt = 1` day) representing the consumption reduction for that day. See [Energy Reduction](#energy-reduction) section for additional information

Other (optional) inputs include:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `EI`         |  Energy intake (kcal) at baseline  | Yes | Model estimate |
|  `fat`        |  Baseline fat mass (kg) of the individual   | Yes | Model estimate |
|  `PAL`        |  Physical activity level           | Yes | $1.5$ |
|  `pcarb_base` |  \% of baseline intake that is carbohydrates | Yes | $0.5$ |
|  `pcarb`      |  \% of intake after day $0$ that is carbohydrates | Yes | `pcarb_base` |

Hence the same female can consume `2000` kcals at baseline, with `22`kg of fat mass, having a physical activity level of `1.7` with a proportion of `0.6` kcals attributed to carbohydrates at baseline and a proportion of `0.5` kcals attributed to carbohydrates after dietary change. 

```{r}
female_model4 <- adult_weight(bw = 80, ht = 1.8, age = 40, sex = "female", 
                      EIchange = rep(-250, 365), EI = 2000, fat = 22, 
                      PAL = 1.7, pcarb_base = 0.6, pcarb = 0.5)
```

Additional inputs include `days` to specify the model's time span and `checkValues` which checks that the model's results are biologically plausible (for example, that modelled body weight is always positive). All inputs used in the model are:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `bw`  |  Body Weight (kg)  | No | - |
|  `ht`  |  Height (m) | No | - |
|  `age` |  Age (yrs) | No | - |
|  `sex` |  Either `'male'` or `'female'` | No | - |
|  `EIchange`  | Energy intake difference from baseline (kcal) | Yes | No energy change |
|  `NAchange`  | Sodium intake difference from baseline (kcal) | Yes | No sodium change |
|  `EI`         |  Energy intake (kcal) at baseline  | Yes | Model estimate |
|  `fat`        |  Fat mass (kg) of the individual   | Yes | Model estimate |
|  `PAL`        |  Physical activity level           | Yes | $1.5$ |
|  `pcarb_base` |  \% of baseline intake that is carbohydrates | Yes | $0.5$ |
|  `pcarb`      |  \% of intake after day $0$ that is carbohydrates | Yes | `pcarb_base` |
| `days`        | Time period (days) to run the model | Yes | $365$ |
| `dt`          | Time step for Runge-Kutta 4        | Yes | $1$ |
| `checkValues` | Check for internal consistency | Yes | `TRUE` |

Finally, we remark that one can also input data from a database to estimate individual-level weight change (see the [related section](#estimating-weight-change-of-a-database)

```{r}
#Database information
weights <- c(45, 67, 58, 92, 81)
heights <- c(1.30, 1.73, 1.77, 1.92, 1.73)
ages    <- c(45, 23, 66, 44, 23)
sexes   <- c("male", "female", "female", "male", "male") 

#Returns a weight change matrix and other matrices
database_model <- adult_weight(weights, heights, ages, sexes)
```

##Plots
Result plots can be obtained by `model_plot` function:

```{r, fig.width=6, fig.height=4}
model_plot(female_model2, "Body_Weight")
```

Plotting options include `"Body_Weight"`, `Fat_Mass`, `Lean_Mass`, `Glycogen`, `Extracellular_Fluid` and `Adaptive_Thermogenesis`. Several can be chosen at the same time:

```{r, fig.width=6, fig.height=4}
model_plot(female_model2, c("Body_Weight", "Fat_Mass"))
```

Variables can also be plotted against age:
```{r, fig.width=6, fig.height=4}
model_plot(female_model2, c("Body_Weight", "Fat_Mass"), timevar = "Age")
```

##Energy Reduction

Energy intake change is usually not continuously measured but measured at different and distant points in time (say 1 year apart). The function `energy_build` allows the user to interpolate different energy models between the interpolation points. As an example consider an individual that by day `365` had reduced his consumption in `250` kcals and by day `730` had increased his consumption by `100` kcals (both compared to the consumption initially). The `energy_build` function interpolates those values via a Brownian Bridge:

```{r}
EIbrownian <- energy_build(c(0, -250, 100), c(0, 365, 730))
```

The interpolation looks like this:

```{r, fig.width=6, fig.height=4}
ggplot() + geom_line(aes(x = 1:730, y = EI), data = data.frame(EI = EIbrownian)) + 
  theme_classic() +
  xlab("Days") + ylab("Energy change (kcals)") + ggtitle("Energy interpolation")
```

Such energy change matrix can be directly inputed in the model:

```{r}
model_brownian <- adult_weight(70, 1.75, 22, "male", EIbrownian, days = 730)
```

Other interpolation modes include `Linear`, `Exponential`, `Stepwise_R` (right stepwise), `Stepwise_L` (left stepwise), and `Logarithmic`:

```{r}
EIlinear      <- energy_build(c(0, -250, 100), c(0, 365, 730), "Linear")
EIexponential <- energy_build(c(0, -250, 100), c(0, 365, 730), "Exponential")
EIstepwise_r  <- energy_build(c(0, -250, 100), c(0, 365, 730), "Stepwise_R")
EIstepwise_l  <- energy_build(c(0, -250, 100), c(0, 365, 730), "Stepwise_L")
EIlogarithmic <- energy_build(c(0, -250, 100), c(0, 365, 730), "Logarithmic")
```

Which look like this:

```{r, fig.width=6, fig.height=4}
ggplot() + 
  geom_line(aes(x = 1:730, y = EI, color = "Brownian"), 
            data = data.frame(EI = EIbrownian)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Linear"), 
            data = data.frame(EI = EIlinear)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Exponential"), 
            data = data.frame(EI = EIexponential)) + 
  geom_step(aes(x = 1:730, y = EI, color = "Right Stepwise"), 
            data = data.frame(EI = EIstepwise_r)) + 
  geom_step(aes(x = 1:730, y = EI, color = "Left Stepwise"), 
            data = data.frame(EI = EIstepwise_l)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Logarithmic"), 
            data = data.frame(EI = EIlogarithmic)) + 
  xlab("Days") + ylab("Energy change (kcals)") + 
  ggtitle("Energy interpolation") +
  theme_classic() + 
  scale_color_manual("Interpolation", 
                     values = c("Brownian" = "red", "Linear" = "deepskyblue3",
                                "Exponential" = "forestgreen", "Logarithmic" = "purple",
                                "Right Stepwise" = "black", "Left Stepwise" = "green"))
```

These models result in different weight changes: 
```{r}
model_linear      <- adult_weight(70, 1.75, 22, "male", EIlinear, days = 730)
model_exponential <- adult_weight(70, 1.75, 22, "male", EIexponential, days = 730)
model_logarithmic <- adult_weight(70, 1.75, 22, "male", EIlogarithmic, days = 730)
model_stepwise_r  <- adult_weight(70, 1.75, 22, "male", EIstepwise_r, days = 730)
model_stepwise_l  <- adult_weight(70, 1.75, 22, "male", EIstepwise_l, days = 730)
```

Which look as follows:
```{r, echo = FALSE, fig.width=6, fig.height=4}
ggplot() +
  geom_line(aes(x = 1:730, y = as.vector(model_linear[["Body_Weight"]]), color = "Linear")) + 
  geom_line(aes(x = 1:730, y = as.vector(model_exponential[["Body_Weight"]]), color = "Exponential")) + 
  geom_line(aes(x = 1:730, y = as.vector(model_logarithmic[["Body_Weight"]]), color = "Logarithmic")) + 
  geom_line(aes(x = 1:730, y = as.vector(model_stepwise_r[["Body_Weight"]]), color = "Right Stepwise")) + 
  geom_line(aes(x = 1:730, y = as.vector(model_stepwise_l[["Body_Weight"]]), color = "Left Stepwise")) + 
  geom_line(aes(x = 1:730, y = as.vector(model_brownian[["Body_Weight"]]), color = "Brownian")) +
  xlab("Days") + ylab("Weight (kg)") + 
  theme_classic() + 
  ggtitle("Weight change under different energy interpolations") + 
    scale_color_manual("Interpolation", 
                     values = c("Brownian" = "red", "Linear" = "deepskyblue3",
                                "Exponential" = "forestgreen", "Logarithmic" = "purple",
                                "Right Stepwise" = "black", "Left Stepwise" = "green"))
```

##Estimating weight change from a database

Vector data can also be used in the model to calculate weight change for several individuals at a time (which is quite faster than doing them individually in a `for` loop). As an example consider the following dataset:

```{r}
#Database information
mydata <- data.frame(
  id     = 1:5,
  weight = c(67, 68, 69, 70, 71),
  height = c(1.30, 1.73, 1.77, 1.92, 1.73),
  age    = c(45, 23, 66, 44, 23),
  sex    = c("male", "female", "female", "male", "male"),
  energy_change = runif(5, -200, 200),
  prob   = c(0.1, 0.2, 0.2, 0.05, 0.45))

#Get energy change with energy build function
eichange      <- energy_build(cbind(rep(0,5), mydata$energy_change), c(0, 365))

#Returns a weight change matrix and other matrices
database_model <- adult_weight(mydata$weight, mydata$height, mydata$age, mydata$sex,
                               eichange)
```

Plots can also be obtained for the population with the same command `model_plot`:

```{r fig.width=6, fig.height=4}
model_plot(database_model, "Body_Weight")
```

Summary measures can be obtained via `model_mean` and `adult_bmi`. The first, quantifies mean for `'Body_Weight'`, `'Adaptive_Thermogenesis'`, `'Extracellular_Fluid'`, `'Age'`, `'Body_Mass_Index'`, `'Glycogen'`, `'Lean_Mass'`, and `'Fat_Mass'`:

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight")
```

```{r, echo = FALSE, warning = FALSE}
head(model_mean(database_model, "Body_Weight"))[,1:5]
```

Mean is only estimated for some points in time, to estimate mean for the whole period, consider changing the `days` vector variable:

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight", days = 1:365)
```

```{r, echo = FALSE, warning=FALSE}
head(model_mean(database_model, "Body_Weight", days = 1:365))[,1:5]
```

Mean can also be grouped by a variable (say, `sex`):

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight", days = 1:365, group = mydata$sex)
```

```{r, echo = FALSE, warning = FALSE}
head(model_mean(database_model, "Body_Weight", days = 1:365, group = mydata$sex))[,1:5]
```

Finally, `model_mean` can also be used to estimate `survey` means using the `svydesign` from the `survey` package:

```{r, eval = FALSE}
require("survey")
design <- svydesign(ids = ~id, probs = ~prob, data = mydata)
model_mean(database_model, group = mydata$sex, design = design)
```

```{r, echo = FALSE, message=FALSE, warning = FALSE}
require("survey")
design <- svydesign(ids = ~id, probs = ~prob, data = mydata)
head(model_mean(database_model, group = mydata$sex, design = design))[,1:5]
```

The same process can be replicated with `adult_bmi` to obtain Body Mass Index prevalence per day:

```{r, eval = FALSE}
adult_bmi(database_model, group = mydata$sex)
```

```{r, echo = FALSE}
head(adult_bmi(database_model, group = mydata$sex))
```

Additional information on the Children model and other package functions can be obtained in the other package's Vignettes

```{r, eval = FALSE}
browseVignettes("bw")
```

#Basic explanation on model's equations {#basic}

\noindent The main idea of the adult dynamic weight change model [@T7, @T8, @T9, @T10] is that body weight $BW(t)$ at a given time $t$ is the sum of four components: lean mass $L(t)$, fat mass $F(t)$, glycogen $g(t)$, and extracellular fluid $ECF(t)$:

\begin{equation}\label{bw}
\overbrace{BW(t)}^{\textrm{Weight}} = \underbrace{F(t)}_{\textrm{Fat}} + \overbrace{L(t)}^{\textrm{Lean}} + \underbrace{3.7 \cdot G(t)}_{\textrm{Glycogen and Associated water}} + \overbrace{ECF(t)}^{\textrm{Extracellular Fluid}}
\end{equation}

Lean mass, fat mass, and glycogen are interrelated, meanwhile the extracellular fluid has its own dynamic. The dynamic model's equations are congruent with the energy conservation law. Changes in weight are therefore determined by:

+ Energy intake ($EI$), 
+ Energy expenditure ($EE$),
+ Glycogen reserve ($G$).


Changes in lean and fat mass are given by changes in intake, expenditure, and glycogen. An increase or decrease in body weight depends on the difference between intake and expenditure. A fraction of the change in weight $p$ will correspond to the change in lean mass and the remaining fraction $(1-p)$ will correspond to changes in fat mass. The corresponding equations are:

\begin{equation}\nonumber
\overbrace{\dfrac{dL}{dt}}^{\textrm{Change in lean mass}} = \underbrace{p}_{\textrm{Fraction}} \overbrace{\frac{\Big( EI - EE - \rho_G \frac{dG}{dt}\Big)}{\rho_L}}^{\textrm{Energy Intake} - \textrm{Expenditure+Storage}}
\end{equation}

\begin{equation}\nonumber
\overbrace{\dfrac{dF}{dt}}^{\textrm{Change in fat mass}} = \underbrace{(1-p)}_{\textrm{Remaining fraction}} \overbrace{\frac{\Big( EI - EE - \rho_G \frac{dG}{dt}\Big)}{\rho_L}}^{\textrm{Energy Intake} - \textrm{Expenditure+Storage}}
\end{equation}

Changes in glycogen are determined by the proportion of energy consumption from carbohydrates $CI$:

\begin{equation}
\overbrace{\dfrac{dG}{dt}}^{\textrm{Change in glycogen}} = \underbrace{\frac{1}{\rho_G} \Big( CI - k_G G^2\Big)}_{\textrm{Carbohydrates and constant parameters}},
\end{equation}

Finally, the equation for extracellular fluid depends on sodium intake and carbohydrates consumed:
\begin{equation}\nonumber
\overbrace{\dfrac{dECF}{dt}}^{\textrm{Change in fluid}} = \overbrace{\frac{ \Big( \Delta Na_{diet} + \xi_{Na}(ECF - ECF_{init}) - \xi_{CI} (1 - \frac{CI}{CI_{b}}) \Big)}{Na}.}^{\textrm{Depends on changes in sodium and carbohydrat intake.}}
\end{equation}

To summarize: weight is divided into four interdependent components: lean mass, fat mass, glycogen, and extracellular fluid. The weight change model describes the processes resulting from interconnections between the components:

\begin{equation}\label{bw2}
\overbrace{BW(t)}^{\textrm{Weight}} = \underbrace{F(t)}_{\textrm{Fat}} + \overbrace{L(t)}^{\textrm{Lean}} + \underbrace{3.7 \cdot G(t)}_{\textrm{Glycogen and Associated water}} + \overbrace{ECF(t)}^{\textrm{Extracellular Fluid}}
\end{equation}


#Advanced explanation on model's equations {#advanced}

The dynamic weight change model @T7 defines individual weight (kg) as the sum of fat $F$ and lean mass $L$, extracellular fluid $ECF$ and glycogen $G$:
\begin{equation}\label{bodyweight}
\begin{aligned}
BW = ECF + 3.7 G + F + L.
\end{aligned}
\end{equation}
Extracellular fluid $ECF \equiv ECF(t)$ is the solution to the ordinary differential equation system: 
\begin{equation}\label{ecf}
\begin{cases}
\dfrac{dECF}{dt} & = \frac{1}{Na} \Big( \Delta Na_{diet} + \xi_{Na}(ECF - ECF_{init}) - \xi_{CI} (1 - CI /CI_{b}) \Big),\\\\
ECF_{init} & = 0.
\end{cases}
\end{equation}
where $Na = 3.22$ mg/ml, $\xi_{Na} = 3000$ mg/L/d, and $\xi_{CI} = 4000$ mg/d, are phisiological constants [@T7].  $\Delta Na_{diet}$ represents the change in sodium (mg/d) for the individual. $CI_b$ is the carbohydrate intake at baseline (assumed to be $1/2$ of energy intake estimated) and $CI \equiv CI(t)$ is the carbohydrate intake after the consumption reduction (assumed to be half the energy intake after reduction, $CI \equiv \Delta EI(t) / 2$). 
Glycogen mass $G \equiv G(t)$ is described by the ordinary differential equation:
\begin{equation}\label{gly}
\dfrac{dG}{dt} = \frac{1}{\rho_G} \Big( CI - k_G \cdot G^2\Big),
\end{equation}
where $\rho_G = 4206.501 \textrm{kcals}/\textrm{kg}$ ($17.6 MJ/kg$), and $k_G = CI_b/G_{init}^2$ is a constant with $G_{init} = 0.5$ kg the initial glycogen mass. 

Fat and lean mass, $F\equiv F(t)$ and $L \equiv L(t)$, represent the solutions to the following system of nonlinear ordinary differential equations: 
\begin{equation}\label{desolveadoras}
    \begin{aligned}
    \dfrac{dF}{dt} & = \frac{\big(1-p \big)}{\rho_F}\Big( EI_{\textrm{init}} - EE - \rho_G \frac{dG}{dt}\Big),\\
    \dfrac{dL}{dt} & = \frac{p}{\rho_L}\Big( EI_{\textrm{init}} - EE - \rho_G \frac{dG}{dt}\Big),
    \end{aligned}
\end{equation}
with $\rho_F = 9440.727$ kcals/kg ($39.5$ MJ/kg), $\rho_L = 1816.444$ kcals/kg ($7.6$ MJ/kg) are constants, and $p \equiv  C/ (C+F)$ a function of fat mass with $C = 10.4 \cdot \rho_L \rho_F^{-1}$. 
Total energy expenditure $EE$ is given by:
\begin{equation}
EE = K + \gamma_F F + \gamma_L L + \delta BW + TEF + AT + \eta_L \dfrac{dL}{dt} + \eta_F \dfrac{dF}{dt} ,
\end{equation}
with $\gamma_F = 3.107075$ kcals/kg/d ($13$ kJ/kg/d), $\gamma_L =  21.98853$ kcals/kg/d ($92$ kJ/kg/d), $\eta_F = 179.2543$ kcals/kg ($750$ kJ/d), $\eta_L =  229.4455$ kcals/kg ($960$ kJ/kg) are physiological constants. $K$ is determined by the initial energy balance condition:
\begin{equation}\label{K}
K =  RMR_{init} \cdot PAL  - \gamma_L L_{init} - \gamma_F F_{init} - \delta BW_{init}
\end{equation}
with $RMR_{init}$ the initial resting metabolic rate, $PAL$ the physical activity level (assumed $PAL = 1.5$), $L_{init}, F_{init}, BW_{init}$ the initial lean, fat and body weight masses. The constant $\delta$ is determined defined as $\delta = RMR_{init} \Big( (1 - \beta_{TEF}) \cdot PAL - 1\Big) / BW_{init}$ with $\beta_{TEF} = 0.1$. Furthermore, the thermal effect of feeding is defined as $TEF \equiv \beta_{TEF} \Delta EI(t)$
with $\Delta EI(t)$ as previously specified. Finally, adaptative thermogenesis is given by the solution to the ODE system:
\begin{equation}\label{at}
\begin{cases}
\dfrac{dAT}{dt} & = \beta_{AT} \Delta EI - AT, \\ \\
AT_{init} & = \beta_{AT} \cdot PAL \cdot RMR_{init}.
\end{cases}
\end{equation}
We remark that for each individual, the initial resting metabolic rate $RMR_{init}$ is described by the equations @mifflin1990new:
\begin{equation}\label{rmr}
RMR_{init} = \begin{cases}
9.99\cdot BW_{init} + 625\cdot H_{init} -4.92\cdot AGE_{init} + 5 &  \text{if Sex $=$ Male},  \\
9.99\cdot BW_{init} + 625\cdot H_{init}-4.92\cdot \cdot AGE_{init} -161 & \text{if Sex $=$ Female}.
\end{cases}
\end{equation}
with $H_{init}$, $AGE_{init}$ initial height and age respectively.
Initial fat mass was obtained via the function:
\begin{equation}\label{fatmass}
F_{init} = \begin{cases}
(1/100) \cdot BW_{init} \cdot \Big(0.14 \cdot AGE_{init} + 37.31 \times \ln(BW_{init} / H_{init}^2) - 103.94 \Big) & \text{if Sex $=$ Male} , \\
(1/100) \cdot BW_{init} \cdot \Big(0.14 \cdot  AGE_{init} + 39.96 \cdot \ln(BW_{init} / H_{init}^2) - 102.01 \Big) & \text{if Sex $=$ Female}.
\end{cases}
\end{equation}

Additional information on the model can be found in [@T7, @T8, @T9, @T10]

## Solution algorithm
To solve this system of differential equations, we used a 4th order Runge-Kutta  algorithm (RK4) @ascher2011first with a stepsize $\Delta t = 1$. RK4 was programmed in `C++` for speed and connected to R via the `Rcpp` package [@Rcpp2, @Rcpp1].

#Additional information
Additional information on the children's model, and why to use the dynamic adult model instead of other classical approaches can be found in the package's vignettes:

```{r, eval = FALSE}
browseVignettes("bw")
```

# References