---
title: "Using the Children Weight Change Model"
author: "Rodrigo Zepeda Tello & Dalia Camacho García Formentí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message=FALSE, warning = FALSE}
require("bw")
require("ggplot2")
```

<!--#The model

The Adult Weight Change Model estimates weight change for adult individuals from the paper in [https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3880593/]. 
-->

#Usage in R

##Inputs
The main inputs for the body weight change model are:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `age` |  Age (yrs) | No | - |
|  `sex` |  Either `'male'` or `'female'` | No | - |

As an example consider a `7` year old `'female'`: 

```{r}
female_model1 <- child_weight(age = 7, sex = "female")
```

Furthermore, the model allows the user to input Fat and Fat Free Mass composition of the body:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `FM`  | Fat Mass (kg) | Yes | Model estimate |
|  `FFM`  | Fat Free Mass (kg) | Yes | Model estimate |

For example, our female might have `19.9` kg of Fat Mass and `5.74` kg of Fat Free Mass:

```{r}
female_model2 <- child_weight(age = 7, sex = "female", FM = 19.9, FFM = 5.74)
```

Energy intake can also be inputed as a vector of daily energy consumption:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `EI`  | Energy Intake per day | Yes | Model estimate |


```{r, eval = FALSE}
female_model3 <- child_weight(age = 7, sex = "female", FM = 19.9, FFM = 5.74,
                              EI = seq(1600, 1750, length.out = 365))
```

Note that in the examples above, `EIchange = seq(1600, 1750, length.out = 365)` is inputed as a vector with each day representing the consumption reduction for that day. See [Energy](#energy) section for additional information.

Other (optional) inputs include:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `days`         |  Number of days to run de model  | Yes | $365$ |
|  `checkValues`  |  Boolean indicating whether the model should check parameters make sense   | Yes | `TRUE` |

All inputs used in the model are:

| Input | Meaning | Optional | Default |
|------:|------:|------:|------:|
|  `age` |  Age (yrs) | No | - |
|  `sex` |  Either `'male'` or `'female'` | No | - |
|  `FM`  | Fat Mass (kg) | Yes | Model estimate |
|  `FFM`  | Fat Free Mass (kg) | Yes | Model estimate |
|  `EI`  | Energy Intake per day | Yes | Model estimate |
| `days`        | Time period (days) to run the model | Yes | $365$ |
| `checkValues` | Check for internal consistency | Yes | `TRUE` |

Finally, we remark that one can also input data from a database to estimate individual-level weight change (see the [related section](#estimating-weight-change-of-a-database)

```{r}
#Database information
ages    <- c(8, 10, 7, 7, 12)
sexes   <- c("male", "female", "female", "male", "male") 

#Returns a weight change matrix and other matrices
database_model <- child_weight(ages, sexes)
```

##Plots
Result plots can be obtained by `model_plot` function:

```{r, fig.width=7, fig.height=4}
model_plot(female_model2, "Body_Weight")
```

Plotting options include `"Body_Weight"`, `Fat_Mass`, `Lean_Mass`, `Glycogen`, `Extracellular_Fluid` and `Adaptive_Thermogenesis`. Several can be chosen at the same time:

```{r, fig.width=7, fig.height=4}
model_plot(female_model2, c("Body_Weight", "Fat_Mass"))
```

Variables can also be plotted against age:
```{r, fig.width=7, fig.height=4}
model_plot(female_model2, c("Body_Weight", "Fat_Mass"), timevar = "Age")
```

##Energy

Energy intake is usually not continuously measured but measured at different and distant points in time (say 1 year apart). The function `energy_build` allows the user to interpolate different energy models between the interpolation points. As an example consider an individual that initially consumed `1600` kcals, by day `365` consumed in `1750` kcals and by day `730` had increased his consumption to `1820` kcals. The `energy_build` function interpolates those values via a Brownian Bridge:

```{r}
EIbrownian <- energy_build(c(1600, 1750, 1820), c(0, 365, 730))
```

The interpolation looks like this:

```{r, fig.width=7, fig.height=4}
ggplot() + geom_line(aes(x = 1:730, y = EI), data = data.frame(EI = EIbrownian)) + 
  xlab("Days") + ylab("Energy intake (kcals)") + ggtitle("Energy interpolation")
```

Such energy change matrix can be directly inputed in the model:

```{r}
model_brownian <- child_weight(10, "male", EI = EIbrownian, days = 730)
```

Other interpolation modes include `Linear`, `Exponential`, `Stepwise_R` (right stepwise), `Stepwise_L` (left stepwise), and `Logarithmic`:

```{r}
EIlinear      <- energy_build(c(1600, 1750, 1820), c(0, 365, 730), "Linear")
EIexponential <- energy_build(c(1600, 1750, 1820), c(0, 365, 730), "Exponential")
EIstepwise_r  <- energy_build(c(1600, 1750, 1820), c(0, 365, 730), "Stepwise_R")
EIstepwise_l  <- energy_build(c(1600, 1750, 1820), c(0, 365, 730), "Stepwise_L")
EIlogarithmic <- energy_build(c(1600, 1750, 1820), c(0, 365, 730), "Logarithmic")
```

Which look like this:

```{r, fig.width=7, fig.height=4}
ggplot() + 
  geom_line(aes(x = 1:730, y = EI, color = "Brownian"), data = data.frame(EI = EIbrownian)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Linear"), data = data.frame(EI = EIlinear)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Exponential"), data = data.frame(EI = EIexponential)) + 
  geom_step(aes(x = 1:730, y = EI, color = "Right Stepwise"), data = data.frame(EI = EIstepwise_r)) + 
  geom_step(aes(x = 1:730, y = EI, color = "Left Stepwise"), data = data.frame(EI = EIstepwise_l)) + 
  geom_line(aes(x = 1:730, y = EI, color = "Logarithmic"), data = data.frame(EI = EIlogarithmic)) + 
  xlab("Days") + ylab("Energy change (kcals)") + ggtitle("Energy interpolation") +
  theme_classic() + 
  scale_color_manual("Interpolation", 
                     values = c("Brownian" = "red", "Linear" = "deepskyblue3",
                                "Exponential" = "forestgreen", "Logarithmic" = "purple",
                                "Right Stepwise" = "black", "Left Stepwise" = "green"))
```

These models result in different weight changes: 
```{r}
model_linear      <- child_weight(10, "male", EI = EIlinear, days = 730)
model_exponential <- child_weight(10, "male", EI = EIexponential, days = 730)
model_logarithmic <- child_weight(10, "male", EI = EIlogarithmic, days = 730)
model_stepwise_r  <- child_weight(10, "male", EI = EIstepwise_r, days = 730)
model_stepwise_l  <- child_weight(10, "male", EI = EIstepwise_l, days = 730)
```

Which look as follows:
```{r, echo = FALSE, fig.width=7, fig.height=4}
ggplot() +
  geom_line(aes(x = 0:730, y = as.vector(model_linear[["Body_Weight"]]), color = "Linear")) + 
  geom_line(aes(x = 0:730, y = as.vector(model_exponential[["Body_Weight"]]), color = "Exponential")) + 
  geom_line(aes(x = 0:730, y = as.vector(model_logarithmic[["Body_Weight"]]), color = "Logarithmic")) + 
  geom_line(aes(x = 0:730, y = as.vector(model_stepwise_r[["Body_Weight"]]), color = "Right Stepwise")) + 
  geom_line(aes(x = 0:730, y = as.vector(model_stepwise_l[["Body_Weight"]]), color = "Left Stepwise")) + 
  geom_line(aes(x = 0:730, y = as.vector(model_brownian[["Body_Weight"]]), color = "Brownian")) +
  xlab("Days") + ylab("Weight (kg)") + 
  ggtitle("Weight change under different energy interpolations") + 
    scale_color_manual("Interpolation", 
                     values = c("Brownian" = "red", "Linear" = "deepskyblue3",
                                "Exponential" = "forestgreen", "Logarithmic" = "purple",
                                "Right Stepwise" = "black", "Left Stepwise" = "green"))
```

##Estimating weight change of a database

Vector data can also be used in the model to calculate weight change for several individuals at a time (which is quite faster than doing them individually). As an example consider the following dataset:

```{r}
#Database information
mydata <- data.frame(
  id = 1:5,
  age = c(8, 10, 7, 7, 12),
  sex = c("male", "female", "female", "male", "male"),
  energy = runif(5, 1500, 2000),
  prob = c(0.1, 0.2, 0.2, 0.05, 0.45))

#Get energy change with energy build function
eichange      <- energy_build(cbind(runif(5, 1500, 2000), mydata$energy), c(0, 365))

#Returns a weight change matrix and other matrices
database_model <- child_weight(mydata$age, mydata$sex, EI = t(eichange))
```

Plots can also be obtained for the population with the same command `model_plot`:

```{r fig.width=7, fig.height=4}
model_plot(database_model, "Body_Weight")
```

Summary measures can be obtained via `model_mean` which quantifies mean for `'Body_Weight'`, `'Fat_Free_Mass'`, and `'Fat_Mass'`:

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight")
```

```{r, echo = FALSE}
head(model_mean(database_model, "Body_Weight"))[,1:5]
```

Mean is only estimated for some points in time, to estimate mean for the whole period, consider changing the `days` vector variable:

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight", days = 1:365)
```

```{r, echo = FALSE, warning=FALSE}
head(model_mean(database_model, "Body_Weight", days = 1:365))[,1:5]
```

Mean can also be grouped by a variable (say, `sex`):

```{r, eval = FALSE}
model_mean(database_model, "Body_Weight", days = 1:365, group = mydata$sex)
```

```{r, echo = FALSE}
head(model_mean(database_model, "Body_Weight", days = 1:365, group = mydata$sex))[,1:5]
```

Finally, `model_mean` can also be used to estimate `survey` means using the `svydesign` from the `survey` package:

```{r, eval = FALSE}
require("survey")
design <- svydesign(ids = ~id, probs = ~prob, data = mydata)
model_mean(database_model, group = mydata$sex, design = design)
```

```{r, echo = FALSE, message=FALSE, warning = FALSE}
require("survey")
design <- svydesign(ids = ~id, probs = ~prob, data = mydata)
head(model_mean(database_model, group = mydata$sex, design = design))[,1:5]
```

Additional information on the model for adults and other package functions can be obtained in the other package's Vignettes

```{r, eval = FALSE}
browseVignettes("bw")
```
